Newly Left-Behind Children's Educational Performance Drops Slightly After Parental Migration

### 2023-02-26 MScThesisText.Rmd
adjusted reference

Globally, millions of children reside in their home communities but live apart from their parents as either one or both parents have migrated for work. These are the so-called "left-behind children" or "stay-behind children." They are common in migrant-sending regions, such as Latin America, Sub-Sahara Africa, East Europe, and large parts of Asia. Researchers estimated that one-sixth of Mexican children [@dewaard2018], one-fifth of Bulgarian children [@romact2021], and one-fourth of Chinese children were separated from their migrating parents [@nbs2017]. The issue of left-behind children has caught the attention of policymakers and legislators. In a resolution, the Council of Europe's @parliamentaryassembly2021 expressed concern about "the scale of this phenomenon and the long-term damage it creates" and declared that leaving "millions of labour migrants' children without parental care is a mass violation of human rights".

@@

Social scientists have researched extensively about China's left-behind children, but the current literature still has three major shortcomings. The first is that most literature focuses on children left behind in rural areas while neglecting the urban counterparts. The National Bureau of Statistics stated that the number of urban left-behind children had been growing and expected that the trend would continue [@nbs2017]. The second shortcoming lies in sampling and research design. Some studies suffer from a limited sample size of less than one thousand participants or a limited geographic coverage of one to a few towns [@liu2021; @yang2022; @shu2021]. Others only estimated correlation using cross-sectional data [@akezhuoli2022; @jin2020; @huang2020]. These studies could have employed causal inference techniques like fixed effect, matching, or inverse probability of treatment weighting. The third shortcoming lies in the measurement of outcome. @chang2019 and @wang2019 measured students' educational performance with the standardized test score of only one academic subject, namely mathematics. To tackle these limitations, I adopted a nationally representative two-wave panel dataset. My sample includes urban left-behind children as well. I included the test results of multiple subjects to measure comprehensive academic achievement.

@@

The other theoretical model is the family disruption model. Children flourish under parental nurturing care. This term refers to the "stable environments that promote health and adequate nutrition, protect from threats ... opportunities for learning," and "responsive, emotionally supportive and developmentally enriching relationships" [@black2021, p. 1]. Parental migration decreases nurturing care, as characterized by less homework supervision and emotional support by parents [@]. Hong talked to a group of left-behind children at a rural school [@hong2019]. These students expressed the utter sense of loneliness "with no advocate, no support and with no one looking out for them" (p. 13). Many of them felt indifferent to grades and disengaged with school. Some even longed to quit school to pursue a free, independent life (p. 14). Similarly, @murphy2022 talked to left-behind adolescents at a village. Some of them "embraced the 'left-behind' label" because it captured their need for more parental care (p. 185), "especially companionship and affection in daily life" (p. 186). According to @chen2023, caregiver-child interactions are infrequent and unstable for left-behind children. She showed that the loss in such interactions "mediates almost all of the adverse effects of parental absence." Her conclusion was based on structural equation modeling (SEM) using a cross-sectional survey with a sample of Chinese children aged 3-15.

@@

I adopted `R` (version 4.2.0) for all the analyses [@R-base]. This involved the following packages: `tidyverse` (version 1.3.2) and `haven` (version 2.5.0) for data cleaning [@R-tidyverse; @R-haven]; `fixest` (version 0.10.4) for econometric modeling [@berge2018; @R-fixest]; `MatchIt` (version 4.4.0) and `cobalt` (version 4.3.2) for matching [@ho2011; @R-MatchIt; @R-cobalt]; `modelsummary` (version 1.0.2), `vtable` (version 1.3.4), and `kableExtra` (version 1.3.4) for presenting the results [@R-modelsummary; @R-vtable; @R-kableExtra]; and `papaja` (version 0.1.1) for typesetting [@R-papaja].

@@

Matching with propensity score does not guarantee a more balanced sample. Suppose that in the matching, some covariates are related the treatment but not to the outcome. In this case, the propensity score estimator's variance would increase. This is like the issue facing linear regression when variables are correlated with the treatment but not with the outcome [@adelson2017; @bergstra2019]. @king2019 argued against PSM in general as it risks more "imbalance, model dependence, and bias" for already balanced data (p. 444). They recommended coarsened exact matching (CEM) as a viable alternative "for data with continuous, discrete, and mixed variables" (p. 450). CEM works better in achieving balance as it optimizes the entire joint distribution of covariates. The method's downside is that it leaves many units without a match and discards these unmatched units [@costalli2021].

@@

This thesis infers the causal relation between parents' migration and their stay-behind children's educational performance in the short run. I build difference-in-differences models with fixed-effect to analyze data from the 2013-2014 and 2014-2015 rounds of China Education Panel Survey. Results suggest that parental migration is, in the short run, negatively associated with children's educational performance. This is evidenced by a 1.2 points drop in academic exam grades and a 2.5 points drop in cognitive test scores. This result contrasts with that of DID models by @bai2018: parental migration is linked to higher English test scores for stay-behind children. The discrepancy could be attributed their study's location, a Northwestern province with grinding poverty. There, the positive impact of resource generation probably outweighs the negative impact of family disruption, resulting in an overall positive effect (p. 1165). In my nationwide sample the reverse may hold. The households may be better off, so labor migration's additional income fails to compensate for the care deficit. Another plausibility is that family disruption precedes resource generation. The former's negative effect befalls the left-behind child soon after parental migration, but the latter's positive effect occurs a while later [@gao2019]. Consequently, I obtain a short-term negative effect.

@@

From my research, we can find new directions for scholarly inquiry. Researchers can explore interaction and moderation in the pathways of resource generation and family disruption. When the CEPS team releases the next round of data, researchers can capitalize a the three-wave panel dataset to build multi-period DID models. This would help examine the parallel trend assumption. Novel methods also offer new possibilities. Machine learning, for example, can estimate the heterogeneous effects across subgroups [@liu2021a]. Looking beyond outbound migration, the return migration of parents also worth studying [@liang2021]. Not many publications have explored return migrants and their formerly left-behind children in China [see @démurger2015; @liu2018].

Let us consider the policy implications. China's multilevel citizenship scheme shall not remain as it is, despite the small reduction of newly left-behind children's schooling performance. The slight drop is only the short-term consequence; The lifelong cumulative effect on children may be extensive [@liang2020; @meng2017]. Besides, educational attainment is only part of human development outcomes. Other outcomes like mental health and non-cognitive skills are also subjected to parental absence's influence [@antia2020]. Policymakers should accelerate the reforms on hukou, opening more doors for parents to migrate together with their children. Besides top-down policymaking, bottom-up approaches are also necessary. School administrators could dedicate more advisory and counseling programs. Social workers could provide paraprofessional training in communities [@wang2020a].

>>>>

Globally, millions of children reside in their home communities but live apart from their parents as either one or both parents have migrated for work. These are the so-called "left-behind children" or "stay-behind children." They are common in migrant-sending regions, such as Latin America, Sub-Sahara Africa, East Europe, and large parts of Asia. Researchers estimated that one-sixth of Mexican children [@dewaard2018], one-fifth of Bulgarian children [@romact2021], and one-fourth of Chinese children were separated from their migrating parents [@nbs2017]. The issue of left-behind children has caught the attention of policymakers and legislators. In a resolution, the Parliamentary Assembly of the Council of Europe expressed concern about "the scale of this phenomenon and the long-term damage it creates" and declared that leaving "millions of labour migrants' children without parental care is a mass violation of human rights" [@pace2021].

@@

Social scientists have researched extensively about the education of Chinese left-behind children, but the current literature still has three major shortcomings. The first shortcoming lies in sampling. For many studies, the geographic coverage is limited to one to a few counties; The participants are children left behind in rural areas but not those in urban areas [see @hu2019; @zhou2014]. @wang2023 noted that the number of urban left-behind children have been growing, and the trend is expected to continue. The second shortcoming is the usage of cross-sectional data [see @huang2022; @liu2022]. @wassink2021 argued that cross-sectional surveys merely focus on single time points, thereby failing to capture the dynamic nature of migratory journey. Research results would be under- or over-estimated "depending on the timing of data collection" (p. 1766). The third shortcoming lies in the measurement of outcome. @chang2019 and @wang2019 measured students' educational performance with the standardized test score of only one academic subject, namely mathematics. To tackle these limitations, I adopted a nationally representative two-wave panel dataset. My sample includes urban left-behind children as well. I included the test results of multiple subjects to measure comprehensive academic achievement.

@@

The other theoretical model is the family disruption model. Children flourish under parental nurturing care. This term refers to the "stable environments that promote health and adequate nutrition, protect from threats ... opportunities for learning," and "responsive, emotionally supportive and developmentally enriching relationships" [@black2021, p. 1]. Parental migration decreases nurturing care, as characterized by less homework supervision and emotional support by parents. Hong talked to a group of left-behind children at a rural school [@hong2019]. These students expressed the utter sense of loneliness "with no advocate, no support and with no one looking out for them" (p. 13). Many of them felt indifferent to grades and disengaged with school. Some even longed to quit school to pursue a free, independent life (p. 14). Similarly, @murphy2022 talked to left-behind adolescents at a village. Some of them "embraced the 'left-behind' label" because it captured their need for more parental care (p. 185), "especially companionship and affection in daily life" (p. 186). According to @chen2023, caregiver-child interactions are infrequent and unstable for left-behind children. She showed that the loss in such interactions "mediates almost all of the adverse effects of parental absence." Her conclusion was based on structural equation modeling using a cross-sectional survey with a sample of Chinese children aged three to fifteen.

@@

I adopted `R` (version 4.2.2) for all the analyses [@R-base]. This involved the following packages: `tidyverse` (version 2.0.0) and `haven` (version 2.5.0) for data cleaning [@R-tidyverse; @R-haven]; `fixest` (version 0.11.1) for econometric modeling [@berge2018; @R-fixest]; `MatchIt` (version 4.5.1) and `cobalt` (version 4.3.2) for matching [@ho2011; @R-MatchIt; @R-cobalt]; `modelsummary` (version 1.3.0), `vtable` (version 1.4.1), and `kableExtra` (version 1.3.4) for presenting the results [@R-modelsummary; @R-vtable; @R-kableExtra]; and `papaja` (version 0.1.1) for typesetting [@R-papaja].

@@

Matching with propensity score does not guarantee a more balanced sample. Suppose that in the matching, some covariates are related the treatment but not to the outcome. In this case, the propensity score estimator's variance would increase. This is like the issue facing linear regression when variables are correlated with the treatment but not with the outcome [@adelson2017]. @king2019 argued against PSM in general as it risks more "imbalance, model dependence, and bias" for already balanced data (p. 444). They recommended coarsened exact matching (CEM) as a viable alternative "for data with continuous, discrete, and mixed variables" (p. 450). CEM works better in achieving balance as it optimizes the entire joint distribution of covariates. The method's downside is that it leaves many units without a match and discards these unmatched units [@costalli2021].

@@

This thesis infers the causal relation between parents' migration and their stay-behind children's educational performance in the short run. I build difference-in-differences models with fixed-effect to analyze data from the 2013-2014 and 2014-2015 rounds of China Education Panel Survey. Results suggest that parental migration is, in the short run, negatively associated with children's educational performance. This is evidenced by a 1.2 points drop in academic exam grades and a 2.5 points drop in cognitive test scores. This result contrasts with that of DID models by @bai2018: parental migration is linked to higher English test scores for stay-behind children. The discrepancy could be attributed their study's location, a Northwestern province with grinding poverty. There, the positive impact of resource generation probably outweighs the negative impact of family disruption, resulting in an overall positive effect (p. 1165). In my nationwide sample the reverse may hold. The households may be better off, so labor migration's additional income fails to compensate for the care deficit. Another plausibility is that family disruption precedes resource generation. The former's negative effect befalls the left-behind child soon after parental migration, but the latter's positive effect occurs some time later [@wassink2021]. Consequently, I obtain a short-term negative effect.

@@

From my research, we can find new directions for scholarly inquiry. Researchers can explore interaction and moderation in the pathways of resource generation and family disruption. When the CEPS team releases the next round of data, researchers can capitalize a the three-wave panel dataset to build multi-period DID models. This would help examine the parallel trend assumption. Novel methods also offer new possibilities. Machine learning, for example, can estimate the heterogeneous effects across subgroups [@liu2021]. Looking beyond outbound migration, the return migration of parents also worth studying [@liang2021]. Not many publications have explored return migrants and their formerly left-behind children in China [see @démurger2015; @liu2018].

Let us consider the policy implications. China's multilevel citizenship scheme shall not remain as it is, despite the small reduction of newly left-behind children's schooling performance. The slight drop is only the short-term consequence; The lifelong cumulative effect on children may be extensive [@liang2020; @meng2017]. Besides, educational attainment is only part of human development outcomes. Other outcomes like mental health and non-cognitive skills are also subjected to parental absence's influence [@antia2020; @pang2022]. Policymakers should accelerate the reforms on hukou, opening more doors for parents to migrate together with their children. Besides top-down policymaking, bottom-up approaches are also necessary. School administrators could dedicate more advisory and counseling programs. Social workers could provide paraprofessional training in communities [@wang2020a].
